---
title: "A subset of an animal survey"
output: html_notebook
---

The data comes from a [short R course](https://datacarpentry.org/R-ecology-lesson) by <https://datacarpentry.org>.

It is survey data of a variety of animal species across different types of plots (control plots, etc).

```{r}
# Read in the original data
surveys <- read.csv("https://ndownloader.figshare.com/files/2292169")
# Show the top six rows
head(surveys)
```

Investigation revealed that the most comment taxon was "Rodent":

```{r}
table(surveys$taxa)
```

First select the rodents from the whole data frame.

```{r}
rodents = subset(surveys, surveys$taxa == 'Rodent')
dim(rodents)
```

Within rodents, there were a few species with relatively large numbers.

```{r}
rodent_counts <- table(rodents$species_id)
rodent_counts
```

We can use the table to find the species id names with more than 2000 individual animals.

```{r}
lots_of_rodents <- rodent_counts[rodent_counts > 2000]
lots_of_rodents
```

Show the total number of rodents from species with > 2000 counts:

```{r}
sum(lots_of_rodents)
```

Get the names of the common species.

```{r}
lots_names <- names(lots_of_rodents)
lots_names
```
 
The `is.element` function tests whether each element of a (say) column is in another list - in our case - the list of species with > 2000 individuals. 
 
```{r}
is_common_rodent <- is.element(rodents$species_id, lots_names)
head(is_common_rodent)
```

Make a new data frame containing only the common rodents:

```{r}
common_rodents <- subset(rodents, is_common_rodent)
head(common_rodents)
```

Show it has the same number of rows as the sum of the counts above:

```{r}
dim(common_rodents)
```

Check that the subset only has the common rodents:

```{r}
table(common_rodents$species_id)
```

Save this data frame as a CSV file, for use in the exercise.

```{r}
write.csv(common_rodents, 'rodent_survey.csv')
```

Read it back in again, to check that it reads correctly:

```{r}
reread_rodents <- read.csv('rodent_survey.csv')
head(reread_rodents)
```

Show the counts of the species:

```{r}
table(reread_rodents$species_id)
```