---
title: "On the t-test and the sampling distribution"
output:
  html_document:
    df_print: paged
---

In the mosquitoes notebook, we saw that the permutation test estimate for the t value was not very like the actual t value.

This will often be so, when the data do not match the assumptions of the t-test.

Here we do a t-test and a permutation test on data that does match the assumptions.

```{r}
# Two groups from the normal distribution
group1 = rnorm(500, mean=10, sd=2)
group2 = rnorm(500, mean=10, sd=2)
```

```{r}
t.test(group1, group2)
```

The observed difference in means:

```{r}
group_mean_diff = mean(group1) - mean(group2)
group_mean_diff
```

The permutation test on our new data.

```{r}
# The permutation test on our new data
pooled = c(group1, group2)
# Make a sequence (vector) of 10000 zeros to store our fake differences
fake_mean_diffs = numeric(10000)
# Repeat 10000 times
for (i in 1:10000) {
  # Repeat the same operations as we did above
  shuffled = sample(pooled)
  fake_1 = head(shuffled, 500)
  fake_2 = tail(shuffled, 500)
  fake_mean_diff = mean(fake_1) - mean(fake_2)
  # Store the fake difference in our list 
  fake_mean_diffs[i] = fake_mean_diff
}
```

The sampling distribution:

```{r}
hist(fake_mean_diffs, main='Sampling distribution of mean difference')
```

The permutation t-a-like:

```{r}
group_mean_diff / sd(fake_mean_diffs)
```

This is similar to the t value we see above.